Coming to Germany: Language, Job search and Labor market outcomes

### 2018-05-18 Paper 2.tex
Update Paper 2.tex

Since the early 1980s, a number of studies has emerged to explore the role of host-country language skills on labor market integration of immigrants. First, language skills, as being a productive trait in itself, have significant direct effect on improving immigrants' labor market outcomes (Yao and van Ours, 2015; Isphording and Otten, 2014; Chiswick and Miller, 2011; Dustmann and Fabbri, 2003). Second, language skills, especially at the beginning of migration, also facilitate early job search and the establishment of social networks (Chiswick and Miller, 2005). Further, many studies in a range of countries find that using personal network of friends and family is a popular and effective search channel to obtain job offers (Wahba and Zenou, 2005; Bentolia et al., 2010; Pellizzari, 2010). But the effectivessness of this search method in terms of later earnings and job quality is questionable as it depends on the quality of an individual's social network composition (Battu et al., 2011; Cappellari and Tatsiramos, 2015).

Empirical evidence in the existing literature on language skills of immigrants often reports the 'total effect' of language fluency on a range of labor market outcomes. But it is, in some cases, not empirically clear \textbf{why} such an effect arises. This study aims to investigate the role of the pre-migration language skills on the use of social networks to find jobs, which in turn affects current earnings and the quality of jobs. In doing so, I attempt to open the black-box to disentangle how and why language proficiency exerts its influence on labor market success of immigrants. I focus on pre-migration language skills because immigrants who master the national language at the very beginning of migration could immediately find better jobs and obtain an initial advantage which puts them well ahead of other immigrants who start learning the language after migration. Further, by focusing on pre-migration language skills, I am able to decompose the total effect of language skills into indirect effect through preferred early job search channels and direct effect on subsequent labor market success.

Traditionally, the mediation analysis is often implemented under the framework of linear structural equation modelling (LSEM) (MacKinnon, 2008). LSEM framework, however, cannot generalize to non-linear relationship in the mediation and outcomes equations (Imai et al., 2010). Hence, this paper implements the potential outcomes framework to accommodate for the binary mediator (using social networks to search for jobs) and the ordered outcome variable (the level of job complexity as a proxy for the quality of jobs).

>>>>

Do people with some host-country language skills at arrival integrate better in the long-run than the ones who do not? Since the 2000s, six Western European countries \footnote{Austria, Denmark, France, Germany, the Netherlands, the UK} have laid down the law that requires pre-departure integration conditions for all non-EU family migrants (Bonjour, 2014). Specifically, since August 2007, foreign adults migrating as spouses of German nationals have to demonstrate at least basic level of German language skills, while foreign children between 16 and 18 years old have to demonstrate advanced level of German language skills, sufficient to start learning in high schools or vocational schools (Block, 2012). If this pre-departure integration requirement makes sense for immigrants coming to Germany for educational or working reasons, does that make equally sense for family reunion purpose?

In response to such kind of questions, a number of studies since the early 1980s has emerged to explore the role of host-country language skills on the integration success of immigrants. Language skills, as being a productive trait in itself, have significant direct effect on improving immigrants' labor market outcomes (Yao and van Ours, 2015; Isphording and Otten, 2014; Chiswick and Miller, 2011; Dustmann and Fabbri, 2003). Further, Berman et al. (2003); Chiswick and Miller (2003) show that destination language fluency also has indirect impacts on wages by complementing other forms of human capital earned in source countries. It follows that people who show readily observable language capability should immediately find better jobs and obtain an initial advantage which puts them well ahead other immigrants who start to learn the language after migration. This is true enough. But is this also true that the admission of family migrants should only be grounded on the moral aspect of family life for which their language skills have no intrinsic value (Bonjour, 2014)?

This study attempts to answer these questions by investigating the role of the pre-migration language skills on the contemporary labor market outcomes of immigrants in Germany, in particular, the earnings and the quality of jobs. However, unlike many existing empirical studies which mainly reports the 'total effect' of language fluency on a range of labor market outcomes, I attempt to open the black-box to disentangle \textbf{how} and \textbf{why} language proficiency exerts its influence on labor market success of immigrants.

Chiswick and Miller (2005) establish that language skills, especially at the beginning of migration, facilitate early job search and the establishment of social networks. At arrival to the host country, immigrants face with various choices of searching for jobs, and they weight different alternatives by their relative costs (Beggs and Chapman, 1990; Frijter et al. 2003; Battu et al., 2011). Immigrants possessing some level of language fluency have lower costs to embrace the mainstream job search methods such as public and private employment services or job advertisements on public media. The ones with poor language skills find it relatively less costly to search jobs via their own informal networks. Many studies in a range of countries find that the job search methods actually matter for the labor market performance. In particular, Wahba and Zenou (2005); Bentolia et al. (2010); Pellizzari (2010) show that using personal network of friends and family is a popular and effective search channel to obtain job offers. But the effectivessness of this search method in terms of later earnings and job quality is questionable as it depends on the quality of an individual's social network composition (Battu et al., 2011; Cappellari and Tatsiramos, 2015).

This paper implements the mediation analysis to decompose the total effect of language skills into indirect effect through preferred early job search channels and direct effect on subsequent labor market success. Traditionally, the mediation analysis is often conducted under the framework of a linear structural equation modelling (LSEM) (MacKinnon, 2008). LSEM framework, however, cannot generalize to non-linear relationship in the mediation and outcomes equations (Imai et al., 2010). Hence, I implement the potential outcomes framework to accommodate for the binary mediator (using social networks to search for jobs) and the ordered outcome variable (the level of job complexity as a proxy for the quality of jobs).
